### 2016-02-02 Projects/DelCo_2012/Report/DelCo_2012.tex
update cafs report and add new references

\headerC{
{
\flushright{
Craig G. Weinschenk \\
Keith M. Stakes \\
Daniel Madrzykowski \\

@@

\headerC{
\flushright{
Craig G. Weinschenk \\
Keith M. Stakes \\
Daniel Madrzykowski \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

\vspace*{2\baselineskip}
August 2015}}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note XXXX \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~XX, \pageref{LastPage} pages (August 2015) \\
% http://dx.doi.org/10.6028/NIST.TN.XXXX \\
CODEN: NTNOEF }

@@

NIST conducted several series of experiments to provide scientific measurements and comparisons between CAFS and water-only hose streams for different compartment fire scenarios.

@@

The key measures were to understand the cooling and suppression capabilities of CAFS compared to water. The experimental designs were developed based on a project planning workshop which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAFS during interior attack operations in residential occupancies~\cite{Grant:2011}.

@@

\subsubsection*{2-Story Structure}

@@

@@

@@

\paragraph{Single-Story Structures}
\label{sec:suppresion_single}

@@

The fuel load for these experiments consisted of one bale of hay, twelve pallets, eleven and a half sheets plywood on the walls and the wood ``floor assembly". Two stacks of pallets with hay were used as the first items ignited in each experiment. Each stack was composed of six pallets with a half bale of hay. The hay was layered between each pallet. Figure~\ref{fig:Wood_Fuel_Load} shows the pallets prior to ignition. The pallets and hay were weighed prior to each experiment. For the four pallet experiments, the average mass of the two stacks of pallets and the bale of hay was 232.4~kg (511.3~lb). A complete listing of the mass of each pallet and hay bales is given in Appendix~\ref{app:fuel_loads}.

\begin{figure}[!ht]
	\includegraphics[width=0.65\columnwidth]{../Figures/Pictures/Wood_Fuel_Package}
	\caption{Photograph of Southeast corner of burn room with wood fuel load.}
	\label{fig:Wood_Fuel_Load}
\end{figure}

@@

\paragraph{Two-Story Structures}
\label{sec:suppresion_two}

@@

@@

\chapter{Fuel Load Details}
\label{app:fuel_loads}

>>>>

\headerC{
{
\flushright{
Craig G. Weinschenk \\
Daniel Madrzykowski \\
Keith M. Stakes \\
Joseph M. Willi \\

@@

\headerC{
\flushright{
Craig G. Weinschenk \\
Daniel Madrzykowski \\
Keith M. Stakes \\
Joseph M. Willi \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

\vspace*{2\baselineskip}
February 2016}}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note XXXX \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~XX, \pageref{LastPage} pages (February 2016) \\
% http://dx.doi.org/10.6028/NIST.TN.XXXX \\
CODEN: NTNOEF }

@@

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but have curtailed the use of these units for this purpose until potential safety implications are better understood. This project is expected to produce a better understanding of the effectiveness and safety implication of CAFS for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAFS for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAFS and plain water to evaluate the effectiveness of CAFS for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAFS and plain water hose lines related to issues such as effective hose stream throw and distribution, forces needed to carry and move hoses streams both horizontally and vertically, and forces need to kink hose lines. The conclusions should include much better information regarding the safe and effective use of CAFS for structural firefighting.

This paper documents Task 3 of the project. NIST conducted several series of experiments to provide scientific measurements and comparisons between CAFS and water-only hose streams for different compartment fire scenarios.

@@

The key measures were to understand the cooling and suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAFS during interior attack operations in residential occupancies~\cite{Grant:2011}. Anecdotally, other fire departments or firefighters had similar incidents or impressions of CAF.

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970’s. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allows the water to penetrate the hot surface of a burning fuel, and also provides additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAFS advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires, but also examined using them for structural fire fighting. Departments across the country began to experiment with CAFS systems for use in structural firefighting. In most documented cases, the findings about the effectiveness of the CAFS were favorable, yet after decades of fire department trials, CAFS still has not been significantly incorporated into structural firefighting tactics. The importance of tactics is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, has now issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations in a structure.

There has never been a comprehensive scientific study on the use of CAFS for structural fire fighting.  Fire departments have conducted pilot programs to examine implementation and effectiveness of CAFS, foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures, and test laboratories and universities have examined small sections of the CAFS puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}. Much of the data has been generated by people and organizations involved in selling the technology to the fire service and, in many cases, the conclusions are based on a very limited set of instruments or comparisons between water and form that were not designed as scientific experiments.

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland firefighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural firefighting, crews typically  have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they may have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior firefighting with CAFS are understood better.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in the State of Maryland and was a committed user of CAFS with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires whiles CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of (MCFRS) issued a General Order prohibiting the use of CAFS during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAFS during structure fires and to report back his findings and recommendations. The catalyst for this study was as a result of this order.

@@

\subsubsection*{Two Story Structure}

@@

\subsubsection*{Single Story Structure}

@@

\subsubsection*{Two Story Structure}

@@

\subsubsection*{Single Story Structure}
\label{sec:suppresion_single}

@@

The fuel load for these experiments consisted of one bale of hay, twelve pallets, eleven and a half sheets plywood on the walls and the wood ``floor assembly". Two stacks of pallets with hay were used as the first items ignited in each experiment. Each stack was composed of six pallets with a half bale of hay. The hay was layered between each pallet. Figure~\ref{fig:Wood_Fuel_Load} shows the pallets prior to ignition. The pallets and hay were weighed prior to each experiment. For the four pallet experiments, the average mass of the two stacks of pallets and the bale of hay was 232.4~kg (511.3~lb).
\begin{figure}[!ht]
	\includegraphics[width=0.65\columnwidth]{../Figures/Pictures/Wood_Fuel_Package}
	\caption{Photograph of Southeast corner of burn room with wood fuel load.}
	\label{fig:Wood_Fuel_Load}
\end{figure}

@@

\subsubsection*{Two Story Structure}
\label{sec:suppresion_two}

@@

In June 2015 two stacks of pallets were added to the fuel load which consisted of three sofas, OSB lining on the ceiling and walls and the polypropylene carpeting over puf padding.  Each stack had 10 pallets.  The mass of each stack was approximately 175 kg .    When hay was added for ignition, the bales of hay averaged 12 kg.

@@
